Tropical Mathematics and the Lambda Calculus

### 2023-01-21 appIV.tex
Commit presentazione del STDLC

Let us end this section by recalling the sound interpretation of $\STDLC$-terms in a $C\partial C$ category (which is taken from [Section 4.3, \cite{Manzo2010}]).

\begin{definition}
Let $\C C$ be a $C\partial C$ and $\model A$ fixed objects of $\C C$, for each ground type $A$ of the $\STDLC$.

@@

Then, one inductively defines an interpretation $\model{\Gamma \vdash M:A}\in\C C(\model \Gamma,\model A)$ of a $\STDLC$-term $\Gamma \vdash M:A$ (that can can be proved to be sound) by induction as:
\begin{itemize}
 \item $\model{\Gamma, x:A \vdash x:A}:=\pi_2$
 \item $\model{\Gamma, x:A \vdash y:B}:=\model{\Gamma \vdash y:B}\circ \pi_1\quad\textit{ if } y\neq x$
 \item $\model{\Gamma\vdash \lambda x.M:A\to B}:=\Lambda(\model{\Gamma,x:A\vdash M:B})$
 \item $\model{\Gamma \vdash M\BB T:B}:=\mathsf{ev} \circ \langle \, \model{\Gamma \vdash M:A\to B}, \model{\Gamma \vdash \BB T:A} \, \rangle$
 \item $\model{\Gamma \vdash \Der [M] \cdot N:A\to B}:=\Lambda\left(\Lambda^-(\model{\Gamma \vdash M:A\to B})\star \model{\Gamma \vdash N:A}\right)$
 \item $\model{\Gamma \vdash 0:A}:=\infty$
 \item $\model{\Gamma \vdash M+\BB T:A}:=\model{\Gamma \vdash M:A}+\model{\Gamma \vdash \BB T:A}$
\end{itemize}
\end{definition}

>>>>

\subsubsection{The language and the semantics of $\STDLC$}

The language we mainly consider in the main paper il the differential $\lambda$-calculus $\STDLC$, that we quickly define here, following [Section 3, \cite{Manzo2010}].

The terms and the sums\footnote{Usually what we call here ``terms'' are called ``simple terms'', and what we call here ``sums'' are called ``differential terms''.} are mutually generated by the following grammars, then quotiented by usual $\alpha$-equivalence plus other equivalences handling the $+$ constructor.

{
\begin{minipage}{\textwidth}
\begin{align*}
\textit{Terms: } & M,N:= x\mid \lambda x.M \mid M\BB T \mid \Der [M] \cdot N
\\
\textit{Sums: } & \BB T:= 0 \mid M \mid M+\BB T
\end{align*}\end{minipage}}\medskip\\
Simple types are generated by the grammar:

{
\begin{minipage}{\textwidth}
\begin{align*}
A,B:= X\mid A\to B
\end{align*}\end{minipage}}\medskip\\
where $X$ varies in a fixed set of ground types.

The typing system is given in Figure~\ref{fig:STDLC}.

\begin{figure}
\fbox{
\begin{minipage}{0.9\textwidth}
\begin{center}

  {
\AXC{$(x: A)\in \Gamma$}
\UIC{$\Gamma\vdash x\in A$}
\DP}

\bigskip

\begin{tabular}{c c }
  {
\AXC{$\Gamma, x: A\vdash M: B$}
\UIC{$\Gamma\vdash \lambda x.M: A\to B$}
\DP}

&

  {
\AXC{$\Gamma \vdash M: A\to B$}
\AXC{$\Gamma\vdash \BB T: A$}
\BIC{$\Gamma \vdash M\BB T: B$}
\DP}

\end{tabular}

\bigskip

  {
\AXC{$\Gamma \vdash M: A\to B$}
\AXC{$\Gamma \vdash N: B$}
\BIC{$\Gamma \vdash \Der [M] \cdot N: A\to B$}
\DP}

\bigskip

\begin{tabular}{c c }

  {
\AXC{}
\UIC{$\Gamma \vdash 0:A$}
\DP}

&

  {
\AXC{$\Gamma\vdash M_1: A \quad \cdots \quad \Gamma \vdash M_n:A \qquad (n\geq2)$}
\UIC{$\Gamma \vdash M_1+\cdots +M_n : A$}
\DP}

\end{tabular}
\end{center}
\end{minipage}
}
\caption{Typing rules for $\STDLC$.}
\label{fig:STDLC}
\end{figure}

We finally recall the sound interpretation of $\STDLC$-terms in a $C\partial C$ category, following [Section 4.3, \cite{Manzo2010}].

\begin{definition}
Let $\C C$ be a $C\partial C$ and $\model X$ fixed objects of $\C C$, one for each ground type $X$ of the $\STDLC$.

@@

Then, one inductively defines an interpretation $\model{\Gamma \vdash M:A}\in\C C(\model \Gamma,\model A)$ of $\Gamma \vdash M:A$ and one of $\model{\Gamma \vdash \BB T:A}\in\C C(\model \Gamma,\model{A})$ of $\Gamma \vdash \BB T:A$ by induction as:
\begin{itemize}
 \item $\model{\Gamma, x:A \vdash x:A}:=\pi_2$
 \item $\model{\Gamma, x:A \vdash y:B}:=\model{\Gamma \vdash y:B}\circ \pi_1\quad\textit{ if } y\neq x$
 \item $\model{\Gamma\vdash \lambda x.M:A\to B}:=\Lambda(\model{\Gamma,x:A\vdash M:B})$
 \item $\model{\Gamma \vdash M\BB T:B}:=\mathsf{ev} \circ \langle \, \model{\Gamma \vdash M:A\to B}, \model{\Gamma \vdash \BB T:A} \, \rangle$
 \item $\model{\Gamma \vdash \Der [M] \cdot N:A\to B}:=\Lambda\left(\Lambda^-(\model{\Gamma \vdash M:A\to B})\star \model{\Gamma \vdash N:A}\right)$
 \item $\model{\Gamma \vdash 0:A}:=:0 \qquad \textit{(the $0$ element of $\C C(\model \Gamma,\model A)$)}$
 \item $\model{\Gamma \vdash M+\BB T:A}:=\model{\Gamma \vdash M:A}+\model{\Gamma \vdash \BB T:A}$ \qquad \textit{(the $+$ operation on $\C C(\model \Gamma,\model A)$)}
\end{itemize}
\end{definition}

One then defines a ``differential'' extension of usual $\beta$-reduction (see again [Section 3, \cite{Manzo2010}], or \cite{ER}), and thanks to the fact that $\C C$ is a $C\partial CC$, the interpretation above can can be proved to be sound for it ([Theorem 4.12, \cite{Manzo2010}]).
